Modelling potential environmental impacts of science activity in Antarctica

### 2025-09-18 paper/paper.qmd
added talk to repo

## Introduction
**FRASER TO FRAME / CLARIFY MOTIVATION**

Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.

Relevant findings as to human impacts in Antarctica. Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.

@@

### GPS data from an expedition
**FRASER**: Timeline, devices used, and associated protocols for scientists while on site.

>>>>

## Introduction
**FRASER MATERIAL STARTS**

Antarctica is often considered a vast, desolate continent devoid of human activity.
However, there are small, scientifically significant areas where substantial, regular human activity occurs.
Globally unique ice-free areas within Antarctica, such as the McMurdo Dry Valleys, are of particular scientific interest and so a significant amount of activity occurs in this specially managed area.
Human activity in Antarctica is highly dependent on the infrastructure that supports people to survive in this environment. 
Outside of a reduced period of activity caused by COVID-19, human activity on the continent continues to grow [@Tejedo2024] and so the actual, and future potential impact of these activities continues to grow. 
With the McMurdo Dry Valleys being very close to the largest research base on the continent, McMurdo Station, the pressures on this unique environment from human activity are increasing. 
Understanding the scope and implications of human activity in key parts of Antarctica is critical to supporting the environmental management of the continent.

The history of understanding the role and impacts of human activity in Antarctica begins with the sixth Antarctic Treaty Consultative Meeting (ATCM) in 1970. 
A recommendation in the final report of this meeting was for member countries to encourage more research on how human activity has impact on Antarctica and its ecosystems [@ATCM1970 Recommendation VI-4].
By the mid-1980s, it was recognised that scientific activity, particularly the logistics required to support that activity, required additional scrutiny, but also that scientific activity could pose a greater threat to Antarctica’s ecosystems than tourism [@benninghoff1985man]. 
At this point, the need for more understanding and the ability to forecast the environmental impact of human activity was recognised [@ATCM1985 Paragraph 46].

Moving into the mid-1990’s, the multi-disciplinary scope of large programmes of research like the McMurdo Dry Valleys Long Term Ecological Research (MCM LTER) programme, increased the number of researchers in the field but also supported more research into the history of human activity in the Dry Valleys [@priscu2016environmental].

With the introduction of the Committee for Environmental Protection (CEP) in 1998, human activities (including science activities) were monitored, but moved away from the understanding and scope of the activities, but more focused on the potential resulting impacts of the activity. 
For example, the current 5-year Work Plan for the CEP (REF), has several items associated with human activities but are focused on managing the impact of human activities rather than quantifying the current level of activity, or forecasting the cumulative impacts of the activity.

More recent research has focused either on a continental-scale picture of human activity (Leihy et al), or detailed mapping of the infrastructural footprint and impacts that the numerous Antarctic stations have on the continent (Brooks et al x 2 Papers). 
These approaches approach the challenge of understanding human activity from contrasting spatial perspectives. 
This research aims to complement these two techniques using high resolution mapping of human activity in the dry valleys to extrapolate the way in which scientists use the landscape to move around.

**FRASER MATERIAL ENDS**

@@

### GPS data from an expedition
**FRASER MATERIAL STARTS**

GPS data showing how scientists moved around the McMurdo Dry Valleys was collected using QStarz Q1100P GPS Tracking Recorders between 2016 and 2018. 
These units were selected because of their long battery life (40+ hours), chipset sensitivity, and simple design and use.

Each unit was set to record a location every 30 seconds. 
Scientists turned on the devices when leaving the base camp in the morning and turned them off when returning at the end of the day. 
The units were attached to the top of a backpack when leaving the camp to ensure the units had bread access to the GPS signal.

Participants were sourced through existing networks of scientists supported by the New Zealand Antarctic Programme who were heading to Antarctica for research purposes. 
All participants were asked if they wish to wear a GPS recorder. 
The study was entirely voluntary. Over the two primary field events, 40 people used the devices. 
Social Ethics approval was received by the Manaaki Whenua Landcare Research Social Ethics committee (SE 1617/05).

![The Qstarz Q1100P GPS Tracking Recorders used to record the movement of research staff while in the McMurdo Dry Valleys.](gps-device-picture.jpg){#fig-gps-device fig-width="60.88mm"}

![30 second locations recorded by ~14 researchers within a single research event within the Miers Valley. The image shows the basecamp, locations of research equipment and/or activities, as well as walking tracks that were taken to access areas further afield.](sample-gps-data.png){#fig-sample-gps-data fig-width="81.28mm"}

**FRASER MATERIAL ENDS**
